---
title: "R Packages"
author: "Lance Santana"
date: "11-01-2024"
format:
  html:
    toc: true
    toc-location: left
    code-line-numbers: true
    embed-resources: true
  pdf:
    code-line-numbers: true
    embed-resources: true
error: true
---

# Installing Packages (not graded)

This assignment and the next few homework assignments require a few packages.
Using `install.packages()`, install the following packages: `maps`, `tidyverse`, and `wordcloud2`.

- If you have issues during installation, check the console for errors.
- Read any error messages from top to bottom.
- If you are unable to finish the assignment due to installation issues, work with a classmate to finish the problem. Fill in all the code chunks as if you were able to use `maps` and `tidyverse` on your computer. *Make a note of the issue in your report right below this bullet point*.

# State Capitals

## Part (a)

Load the `maps` and `tidyverse` packages with `library()`.

```{r}
library(maps)
library(tidyverse)
```

**Do you see any conflicts in the tidyverse report?**
No
::: callout-tip
## Documentation

Remember that you can look up information about a package using `help(package = "<the package name>")`.

:::

## Part (b)

Create a plot with the 48 continental US states using `map("state")` from the `maps` package. In the code chunk, add the following options:

- `fig-align: center`
- `width: 80%`

::: callout-tip
## Hint
If you're having trouble, go back and study the output in Part (a).
:::

```{r, fig.align='center',width='80%'}
maps ::map("state")
```


## Part (c)

Recall that the `maps` package has a database on US cities, called `us.cities`.

- Convert that object (`data.frame`) into a `tibble`, calling the new object `us.capitals`.
- Display the first 5 rows of `us.capitals`.

```{r}
us.capitals = as_tibble(us.cities)
head(us.capitals, 5)
```

## Part (d)

Using the `filter()` function on `us.capitals`, select only the **capital** cities in the **continental US** *and* modify the variable. Display the first 10 rows.

::: callout-tip
### Hint

- Note that capital cities are indicated with `capital == 2`.
- There is more than one way to do this problem!

:::

```{r}
us.capitals = us.capitals %>% filter(capital == 2 & long > -125 & long < -66.9)
head(us.capitals, 10)
```


## Part (e)

Plot the continental US again and add `points()` indicating capital cities using the `long` and `lat` variables.

```{r}
maps ::map("state")
points(us.capitals$long, us.capitals$lat, col = 'red', pch = 8)
```

