---
title: "String Manipulation"
author: "Lance Santana"
date: "11-08-2024"
format:
  html:
    toc: true
    toc-location: left
    code-line-numbers: true
    embed-resources: true
  pdf:
    code-line-numbers: true
    embed-resources: true
error: true
---

This exercise uses the following set of lyrics. Each line is stored as a `character` object. **Do not modify it**.

```{r}
lyrics <- c(
  "The monkeys stand for honesty",
  "Giraffes are insincere",
  "And the elephants are kindly, but they’re dumb",
  "Orangutans are skeptical",
  "Of changes in their cages",
  "And the zookeeper is very fond of rum",
  "Zebras are reactionaries",
  "Antelopes are missionaries",
  "Pigeons plot in secrecy",
  "And hamsters turn on frequently",
  "What a gas!",
  "You gotta come and see",
  "At the zoo",
  "At the zoo"
)
```


In the *second* line, use a function to **subset** the string starting at position 14 and ending at position 23.

```{r}
# your code
library(tidyverse)
lyrics[2]
substr(lyrics[2], 14, 23)
```

Use a function to **split** all the lyrics and extract all the words. Convert the result into a `character` vector (that is, *not* a list object).

```{r}
# your code
extracted_vector = str_split(lyrics, pattern = " ")
chr_vector = unlist(extracted_vector)
chr_vector
```

Use a function to **join** the first three lines using a newline symbol. Display the resulting string using `cat()`. *There should be no extra spaces in the result*.

```{r}
# your code
cat(paste(lyrics[1:3],collapse = '\n'))
```

Use a function to **substitute** the word `"rum"` with `"Rum"` in line 6.

```{r}
# your code
replaced_string = str_replace(lyrics[6], "rum", "Rum")
replaced_string
```

Use a function to **search** for all the lines that contain the sequence `"zoo"`.

```{r}
# your code
find_string = str_detect(lyrics, 'zoo')
search_string = lyrics[find_string]
search_string
```
